Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-05-21 04-analisis.qmd
Up graficos interaccion, tablas, citas y material suplementario + render

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
---

@@

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               influence.ME, 
               marginaleffects,
               MLMusingR,
               texreg, 
               ggdist,
               misty,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools)

@@

datos.health <- df_study1 %>% 
  mutate(just_health = factor(just_health, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>% 
  group_by(idencuesta, ola) %>% 
  count(just_health) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                      "2017",
                                      "2018",
                                      "2019",
                                      "2022",
                                      "2023")))

etiquetas.health <- df_study1 %>%
  mutate(just_health = factor(just_health, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>% 
  group_by(ola, just_health) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

datos.pension <- df_study1 %>% 
   mutate(just_pension = factor(just_pension, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>% 
  group_by(idencuesta, ola) %>% 
  count(just_pension) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.pension <- df_study1 %>%
  mutate(just_pension = factor(just_pension, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>% 
  group_by(ola, just_pension) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

datos.educ <- df_study1 %>% 
  mutate(just_educ = factor(just_educ, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>%
  group_by(idencuesta, ola) %>% 
  count(just_educ) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.educ <- df_study1 %>%
    mutate(just_educ = factor(just_educ, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor disagree",
                                         "Disagree",
                                         "Strongly disagree"))) %>%
  group_by(ola, just_educ) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

```{r}
interactions::interact_plot(m13.2, pred = "merit_effort_mean", modx="perc_inequality_mean", mod2="ola_num")
```

@@

texreg::knitreg(list(m1, m2, m3, m4, m5, m6, m7),
               custom.model.names = c(paste0("Model ", seq(1:7))),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               omit.coef = omit,
               custom.coef.map = ccoef,
               digits = 3,
               groups = list("Wave (Ref.= 2016)" = 2:6),
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.79,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.rows = list("Controls"=c(rep("No",6), "Yes")))

@@

texreg::knitreg(list(m8.1,m8.2,m9.1,m9.2),
               custom.model.names = c("Model 8", "Model 9", "Model 10", "Model 11"),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.8,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.rows = list("Controls"=c(rep("Yes",4))))

@@

```{r}
#| label: fig-interact
#| out-width: '100%'
#| fig-asp: 1
#| fig-cap: "Within effects of perception of inequality and meritocratic perceptions on market justice preferences"
#| fig-cap-location: top
#| echo: false
#| results: asis

@@

p6 <- ggpredict(m9.1, terms = c("merit_effort_mean", "perc_inequality_mean")) %>% 
  plot() + 
 # scale_x_continuous(n.breaks = 6) +
  scale_color_manual(
    name = "Perception\nInequality (BE)",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")
  ) +
  scale_fill_manual(
    name = "",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")) +
  labs(x = "Merit: Effort (BE)",
       title = "(A) People are rewarded for their efforts",
       color = NULL) +
  ggdist::theme_ggdist() +
  theme(legend.position = "bottom")

p7 <- ggpredict(m9.2, terms = c("merit_talent_mean", "perc_inequality_mean")) %>% 
  plot() +
#  scale_x_continuous(n.breaks = 6) +
  scale_color_manual(
    name = "Perception\nInequality (BE)",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")
  ) +
  scale_fill_manual(
    name = "",
    values = c("#D55E00", "#0072B2", "#009E73"),
    labels = c("-1 SD", "Mean", "+1 SD")
  ) +
  labs(
    x = "Merit: Talent (BE)",
    title = "(B) People are rewarded for their intelligence",
    caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)"
  ) +
  ggdist::theme_ggdist() +
  theme(legend.position = "bottom")

ggpubr::ggarrange(p6, p7, common.legend=TRUE, ncol=1)

@@

@tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic perceptions on market-justice preferences. Contrary to our hypothesis, the interaction terms in the within-person specification (Model 8 and Model 9) do not reach statistical significance. However, in the between-person specification (Model 10 and Model 11), interaction terms reach statistical significance, indicating that perceptions of economic inequality do meaningfully alter the effect of meritocracy on support for market allocation of social services. These effects are positive and mean that when the perception of inequality is greater, the positive effect of the meritocratic variables on market justice preferences is greater. These interaction effects are visualized in @fig-interact

@@

texreg::knitreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
               custom.model.names = c("Model 12", "Model 13", "Model 14", "Model 15", "Model 16", "Model 17"),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.8,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.rows = list("Controls"=c(rep("Yes",6))))

@@

```{r}
#| label: fig-growth
#| out-width: '100%'
#| fig-asp: 1
#| fig-cap: "Time varying within effects of meritocratic perceptions on market justice preferences"
#| fig-cap-location: top
#| echo: false
#| results: asis

@@

With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time only has significant effects on the between effects of the meritocratic variables (Model 16 and Model 17). These interaction effects are negative and mean that in the last waves, the impact of meritocratic variables on market justice preferences is attenuated. @fig-growth illustrates the interaction between survey wave and the within-person effects of the meritocracy variables on market-justice preferences. As time advances, the positive within-person effect of perceiving that effort is rewarded (panel A) or that intelligence is rewarded (panel B) progressively attenuates. Put differently, the uplift in market justice preferences associated with a strong meritocratic perception becomes smaller in each successive wave.

>>>>

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
editor_options: 
  chunk_output_type: console
---

@@

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               influence.ME, 
               marginaleffects,
               MLMusingR,
               texreg, 
               ggdist,
               misty,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools,
               interactions)

@@

datos.health <- df_study1 %>% 
  mutate(just_health = factor(just_health, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"),
                              ordered = T)) %>% 
  group_by(idencuesta, ola) %>% 
  count(just_health) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                      "2017",
                                      "2018",
                                      "2019",
                                      "2022",
                                      "2023")))

etiquetas.health <- df_study1 %>%
  mutate(just_health = factor(just_health, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"),
                              ordered = T)) %>% 
  group_by(ola, just_health) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

datos.pension <- df_study1 %>% 
   mutate(just_pension = factor(just_pension, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"),
                              ordered = T)) %>% 
  group_by(idencuesta, ola) %>% 
  count(just_pension) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.pension <- df_study1 %>%
  mutate(just_pension = factor(just_pension, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"),
                              ordered = T)) %>% 
  group_by(ola, just_pension) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

datos.educ <- df_study1 %>% 
  mutate(just_educ = factor(just_educ, 
                            levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"),
                              ordered = T)) %>%
  group_by(idencuesta, ola) %>% 
  count(just_educ) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.educ <- df_study1 %>%
    mutate(just_educ = factor(just_educ, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"),
                              ordered = T)) %>%
  group_by(ola, just_educ) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 5 ~ "2022",
                          ola == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

@@

```{r, include=FALSE, eval=FALSE}
interactions::interact_plot(m13.2, pred = "merit_effort_mean", modx="perc_inequality_mean", mod2="ola_num")
```

@@

texreg::knitreg(list(m1, m2, m3, m4, m5, m6, m7),
               custom.model.names = c(paste0("Model ", seq(1:7))),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               omit.coef = omit,
               custom.coef.map = ccoef,
               digits = 3,
               groups = list("Wave (Ref.= 2016)" = 2:6),
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.79,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.names = c("BIC", "Numb. obs.", "Num. groups: individuals", "Var: individuals (Intercept)", "Var: Residual", "Var: individuals, wave", "Cov: individuals (Intercept), wave"),
               custom.gof.rows = list("Controls"=c(rep("No",6), "Yes")))

@@

texreg::knitreg(list(m8.1,m8.2,m9.1,m9.2),
               custom.model.names = c("Model 8", "Model 9", "Model 10", "Model 11"),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.8,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.names = c("BIC", 
                                    "Numb. obs.", 
                                    "Num. groups: individuals", 
                                    "Var: individuals (Intercept)", 
                                    "Var: individuals, merit effort cwc", 
                                    "Var: individuals, perception inequality cwc", 
                                    "Cov: individuals (Intercept), merit effort cwc",
                                    "Cov: individuals (Intercept), perception inequality cwc",
                                    "Cov: individuals, merit effort cwc, perception inequality cwc",
                                    "Var: Residuals",
                                    "Var: individuals, merit talent cwc", 
                                    "Cov: individuals (Intercept), merit talent cwc",
                                    "Cov: individuals, merit talent cwc, perception inequality cwc"),
               custom.gof.rows = list("Controls"=c(rep("Yes",4))))

@@

```{r}
#| label: fig-interact
#| out-width: '100%'
#| fig-asp: 1
#| fig-cap: "Between effects of meritocratic perceptions on market justice preferences by economic inequality perception"
#| fig-cap-location: top
#| echo: false
#| results: asis

@@

p6 <- marginaleffects::plot_slopes(model = m9.1,
                             variables = "merit_effort_mean",
                             condition = "perc_inequality_mean",
                             conf_level = .95) +
  geom_hline(yintercept = 0,
             color = "red",
             linetype = "dashed") +
  labs(x = "Perception inequality (BE)",
       y = "Marginal effect of Merit: Effort (BE)",
       fill = NULL,
       title = "(A) People are rewarded by their efforts") +
  theme_ggdist()

p7 <- marginaleffects::plot_slopes(model = m9.2,
                             variables = "merit_talent_mean",
                             condition = "perc_inequality_mean",
                             conf_level = .95) +
  geom_hline(yintercept = 0,
             color = "red",
             linetype = "dashed") +
  labs(x = "Perception inequality (BE)",
       y = "Marginal effect of Merit: Talent (BE)",
       fill = NULL,
       title = "(B) People are rewarded for their intelligence",
       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)
       Confidence intervals at 95%") +
  theme_ggdist()

library(ggpubr)
ggpubr::ggarrange(p6, p7, common.legend=TRUE, ncol=1)

@@

@tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic beliefs on market justice preferences. Contrary to our hypothesis, the interaction terms in the within-person specification (Models 8 and 9) do not reach statistical significance. However, in the between-person specification (Models 10 and 11), the interaction terms are statistically significant, suggesting that perceptions of economic inequality meaningfully alter the effect of meritocracy on support for market-based allocation of social services. These positive effects indicate that, as perceptions of inequality increase, the positive association between meritocratic perceptions and market justice preferences becomes stronger. The marginal effects of this interaction are illustrated in @fig-interact, which shows that the effect of meritocratic perceptions on market justice preferences intensifies as perceived economic inequality rises.

@@

texreg::knitreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
               custom.model.names = c("Model 12", "Model 13", "Model 14", "Model 15", "Model 16", "Model 17"),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
               leading.zero = T,
               use.packages = F,
               booktabs = F,
               scalebox = 0.8,
               include.loglik = FALSE,
               include.aic = FALSE,
               center = T,
               custom.gof.names = c("BIC", 
                                    "Numb. obs.", 
                                    "Num. groups: individuals", 
                                    "Var: individuals (Intercept)",

                                    "Var: individuals, perception inequality cwc",
                                    "Var: individuals, wave",
                                    "Cov: individuals (Intercept), perception inequality cwc",
                                    "Cov: individuals (Intercept), wave",
                                    "Cov: individuals, perception inequality cwc, wave",
                                    "Var: Residuals",
                                    "Var: individuals, merit effort cwc", 
                                    "Cov: individuals (Intercept), merit effort cwc",
                                    "Cov: individuals, merit effort cwc, wave",

                                    "Var: individuals, merit talent cwc", 
                                    "Cov: individuals (Intercept), merit talent cwc",
                                    "Cov: individuals, merit talent cwc, wave",

                                    "Var: individuals, perception inequality mean", 
                                    "Cov: individuals (Intercept), perception inequality mean",
                                    "Cov: individuals, perception inequality mean, wave",

                                     "Var: individuals, merit effort mean", 
                                    "Cov: individuals (Intercept), merit effort mean",
                                    "Cov: individuals, merit effort mean, wave",

                                    "Var: individuals, merit talent mean", 
                                    "Cov: individuals (Intercept), merit talent mean",
                                    "Cov: individuals, merit talent mean, wave"
                                    ),
               custom.gof.rows = list("Controls"=c(rep("Yes",6))))

@@

```{r}
#| label: fig-growth
#| out-width: '100%'
#| fig-asp: 1
#| fig-cap: "Time varying within effects of meritocratic perceptions on market justice preferences"
#| fig-cap-location: top
#| echo: false
#| results: asis
#| include: false
#| eval: false

@@

With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time only has significant effects on the between effects of the meritocratic variables. These interaction effects are negative and mean that in the last waves, the impact of meritocratic variables on market justice preferences is attenuated. Specifically, as time advances, the positive within-person effect of perceiving that effort is rewarded (Model 16) or that intelligence is rewarded (Model 17) progressively attenuates. Put differently, the uplift in market justice preferences associated with a strong meritocratic perception becomes smaller in each successive wave.

### 2025-05-21 05-discussion.qmd
Up graficos interaccion, tablas, citas y material suplementario + render

The first hypothesis suggested that perceptions of economic inequality were strongly linked to attitudes toward market justice. The findings demonstrated that individuals perceiving higher income disparities showed stronger support for the distribution of social services based on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies (Castillo et al., 2012; Mijs, 2021). This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity. Interestingly, the negative within-effects observed over time implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system. This indicated that while perception heightened awareness, prolonged exposure or social mobilization could also foster skepticism about whether redistribution effectively addressed inequality or simply perpetuated systemic flaws.

Regarding the second hypothesis, the results reinforced that meritocratic perceptions ---particularly those emphasizing individual effort and talent--- were associated with stronger support for market-based distribution systems. Individuals who viewed success as primarily earned through effort were more inclined to justify existing inequalities, seeing them as outcomes of individual merit rather than systemic injustice. This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification (Castillo et al., 2012; Hoyt et al., 2023). These beliefs functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved. The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes.

>>>>

The first hypothesis suggested that perceptions of economic inequality were strongly linked to attitudes toward market justice. The findings demonstrated that individuals perceiving higher income disparities showed stronger support for the distribution of social services based on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021]. This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity. Interestingly, the negative within-effects observed over time implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system. This indicated that while perception heightened awareness, prolonged exposure or social mobilization could also foster skepticism about whether redistribution effectively addressed inequality or simply perpetuated systemic flaws.

Regarding the second hypothesis, the results reinforced that meritocratic perceptions ---particularly those emphasizing individual effort and talent--- were associated with stronger support for market-based distribution systems. Individuals who viewed success as primarily earned through effort were more inclined to justify existing inequalities, seeing them as outcomes of individual merit rather than systemic injustice. This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification [@Castillo2012a_justice; @hoyt_mindsets_2023]. These beliefs functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved. The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes [@canalesceron_sujeto_2021].
